Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-20 paper.tex

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: the
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce
  new therapeutic molecules.  Only within the last decade -- as an
  effect of a change in mindsets, government-mandated policy changes,
  and, not least, as a result of chemists volunteering their time
  for an Open Science ``movement'' -- have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
\end{abstract}

@@

Novel technologies in the life sciences churn out information at an
ever increasing rate -- public data stores such as that of the
European Bioinformatics Institute (EBI) contain in the order of 10
petabytes of biological information.  While biological information has
been publicly available for decades, the same has not been true for
chemical information, until recently. It was not until 2004 that a
large public small molecule structure repository (PubChem) became
freely available. This was followed by other databases. While much of
the foundational algorithms of cheminformatics have been described
since the 1950s, publicly available (open source) software implementing many of
these algorithms has only been accessible in the last ten to fifteen
years \cite{faulon2010}.

@@

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (A), a labeled graph (B), a 3D
  conformation (C) and a surface represention (D). All show differing
  (and valid) aspects of the same molecule and are suited for
  different purposes. The choice of representation is guided by the
  intended use of the structure data.}
\label{figure:chemical-structures}
\end{figure}

@@

Even though some structure representations contain more explicit
information than others, they are all equally valid and are suited for
different problems. Indeed, Fig.~\ref{figure:chemical-structures}
represents a hierarchy of representations. Thus, when searching for
substructures, a 2D representation is sufficient. When trying to
rationalize why a molecule binds to a protein, a 3D structure is
vital.

@@

There is considerable progress in open chemical and bioactivity
databases like ChEMBL and PubChem, which are both freely
available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points each. Integration of these disparate
data is essential for researchers to obtain the fullest possible 
perspective on what is presently known, tracking ongoing advances in 
science as these become available. Still, the integration of
chemical databases with one another remains a challenge, not only due
to the normalization problems in the chemical and bioactivity
landscape, but also due to sheer data volume. One problem is the
annotation of the ``protein targets'' a molecule binds to, as  
dozens of protein and gene identifiers exist, generally without the
availability of a one-to-one mapping from each ontology, which makes
merging databases often a cumbersome exercise.
% [JH comment] 
% Is this above sentence really true? I thought there was general standardization on 
% UniProt for proteins, GenBank for genes, and that mappings between the other
% sorts of identifiers to these primary ones were actually well maintained?
% As I understand it, the main problem with target representation is that
% protein/gene targets are often unknown, and then it is necessary to map to
% pathway, cell, tissue, biological process etc. --> representing the maximum
% that is known about the mechanism of action for the molecule -- and THESE
% data types are often not well standardized
% [JH comment]

Given the requirement to ``make sense'' of as much chemical data at
hand as possible, we have to be able to search in complex data, for
example multi-labelled molecular graphs, where graph labels can change
from database to database. Hence, we need to be able to capture
molecular graphs in a risk-bounded machine encoded way. %risk-bounded meaning here not clear

A fundamental challenge in cheminformatics is the unique
representation of chemical structures in a machine readable format. A
(chemical) graph can be encoded in multiple ways, depending on how one
orders the nodes, resulting in multiple representations of the same
molecule. When one considers the need to include other aspects such as
protonation states (how the molecule changes when one proton is added
to it) and tautomer states (how the molecule can change when a proton
migrates from one part of the molecule to another), the challenge  %This is repeating something stated earlier
increases. The need for a unique representation that is invariant with
respect to the atom ordering arises due to the fact that graph
isomorphism (i.e., checking whether two structures represent the same molecule) is an
expensive operation. The first such ``unique representation
algorithm,'' or \emph{canonicalization} algorithm, was described by
Morgan \cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format \cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

@@

\subsection{Situating Molecular Data in Biological Context}
\label{sec:prof-ident}

The preceding section focuses on the choice of representation of
chemical structures. In the quest for the discovery of novel
therapeutic agents, it is essential to study molecules as they are
situated in a biological context, underlying the processes of life, 
as metabolites, cofactors for enzymatic reactions, hormones, signalling
molecules and environmental xenobiotics. This gives rise to the need
to adequately represent and organise chemical data across biological
databases such as those containing pathways, protein information or 
biological samples. Integration and processing of data from such 
disparate domains underlies ``systems'' level biological research.
This gives rise to some additional challenges that standard chemical
structural representation cannot address: firstly, the representation of
\textit{classes of chemical entities} -- since, in many cases, a class of 
compounds behave in a certain context in a biological system rather than
a single molecule with a fully-specified structure -- and for similar
reasons, chemical concepts such as \textit{groups} (parts of molecules); 
secondly, the need to represent \textit{non-structural} groupings of 
compounds of interest, since compounds may bind to the same or similar 
targets or they may have  similar biological functions such as acting as 
an anti-neoplastic -- preventing the development of tumors. With the 
many different biological databases needing to refer to these different
chemical entities in an organised and standardized 
fashion to enable cross-database integration for a whole-systems 
perspective, \textit{ontologies} and other semantic technologies 
are being successfully employed to support annotation, classification and
semantic cross-domain querying of chemical entities within
biological systems. The most widely used ontology for biologically relevant
chemical entities is ChEBI \cite{chebi2010}, containing around 26,500 
entities in an interlinked semantic graph as of November 2011, and
being actively used for annotation in dozens of biological databases. 
Ontologies are based on logical languages of 
varying levels of expressivity, accompanied by sophisticated reasoning
algorithms. They can be used to construct various sorts of semantic
similarity (e.g. similarity in function, similarity in application) 
which complement traditional structural similarity methods \cite{couto2010}.
An open challenge is integrating graph-based chemical
structure information with the wider logic-based ontology information, which 
will allow combined automated reasoning and querying over the different domains 
encoded in a growing number of interlinked ontologies \cite{hastingsowled2010}. 
Enabling automated inferencing in such heterogenous linked systems brings us closer to
achieving a systems level understanding of bioactive chemistry \cite{Oprea:2007fk}.

@@

While the complexity of molecular graph enumeration remains an open
problem, 
% [JH comment]
% Not sure what is meant here? Surely, the complexity of molecular graph enumeration 
% is upper-bounded exponentially in the number of atoms? Additional complexity
% arises from algorithms to select "good" candidates from the candidates?
% Perhaps what is meant is not that the complexity itself is an open problem,
% but finding methods for enumeration which are more efficient is an open problem?
% Alternatively, if complexity in the CS sense is not what is meant here, maybe it will
% help this audience if a different word is used. 
% [JH comment]
an alternative to enumerating the space is to sample it
\cite{goldberg1999}. Sampling procedures based on the Metropolis or
Genetic Algorithms have been developed and used to elucidate
compounds from NMR data. It should be noted that even in the
face of the computational complexity of molecular enumeration,
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB} (but only
considering molecules up to 13 heavy atoms and using only carbon,
nitrogen, oxygen, sulfur and chlorine). Yet, the current enumeration
software products do not generally produce stereoisomers or tautomers;
these require specific enumeration procedures, which are still the
subject of research by the cheminformatics community.

Another relevant application of enumeration methods is in the
\emph{generation of chemical reaction networks}. The problem here
consists of enumerating all the possible compounds that can be
produced by applying reaction rules to a set of initial molecules. By
reversing the reaction rules one can also find sets of starting
compounds necessary for producing a given target, known as
\emph{retrosynthesis}. Designing new drugs or chemicals, understanding
the kinetics of combustion and petroleum refining, studying the
dynamics of metabolic networks, applying metabolic engineering and
synthetic biology to produce heterologous compounds (compounds from
different species) in microorganisms, all involve enumeration of
reaction networks. As reviewed in Chapter 11 in \cite{faulon2010}
several network enumeration techniques have been developed. However,
these techniques generally suffer from a combinatorial explosion of
product compounds. One way to limit the number of compounds generated
is to simulate the dynamics of the network while it is being
constructed and remove compounds of low concentration. Following that
idea, methods have been developed based on the Gillespie Stochastic
Simulation Algorithm (SSA) to compute on-the-fly species
concentrations. Chemical reaction network enumeration and sampling is
an active field of research, particularly in the context of
metabolism, either to study biodegradation, or to propose metabolic
engineering strategies to biosynthesize compounds of commercial
interest. The difficulty with metabolic network design is that in
addition to network generation based on reactions, one also needs to
verify that there are enzymatic events possible for enabling reactions
catalyses (in other words, enzymes must be present to reduce the
energy required for some reactions to take place). That additional
task requires encompassing both chemical structure and protein
sequence information and the development of tools that are at the
interface between cheminformatics and bioinformatics.

@@

The basis of predictive modeling in cheminformatics is that the
biological activity of a molecule is a function of its chemical
structure. Together with the \emph{similar property principle}
\cite{Johnson:1990qf} (structurally similar molecules will exhibit
similar properties), the goal of any modeling approach is to capture
and characterize the correlations between structural features and the
observed biological activities. Simultaneously, such approaches must
also describe the likelihood of error when using such
models for decision making.  A variety of approaches can be employed
to assess the error in (or conversely, the reliability or confidence of) a
prediction, ranging from statistical approaches to more
empirical approaches such as defining an \emph{applicability domain} -
the region of the input (i.e., chemical structures) that can be
reliably predicted, usually deliniated using some form of similarity
to the training set.

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
  Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking (which predicts how molecules fit together),
pharmacophore modeling (a 3D approach to capturing protein-ligand
interactions) and proteochemometric methods have been designed to take
into account both ligand and receptor structures. The last method is
an example of an extension of statistical QSAR methods to simultaneous
model the receptor and ligand as first reported by Lapinsh \textit{et
  al.}~\cite{lapinsh2001}.

@@

Both methods have advantages and disadvantages. The vectorial approach
requires one to identify a subset of \emph{relevant} (to the property
being modeled) descriptors. This is the feature selection problem and
is well discussed in the data mining literature. A kernel approach
does not require feature selection; but now, one faces the problem of
evaluating a data set in a pairwise fashion. Furthermore, one must now
identify an appropriate kernel. This is an important challenge as the
kernel should be selected to satisfy \emph{Mercer's condition} (a
well-known mathematical property in machine learning that makes a set
of observations easier to make predictions about), and this is not
always possible with traditional cheminformatics based kernels (such
as a multiple common substructure based kernel). These challenges can
make kernel based methods prohibitive on very large datasets.

@@

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in isolation. 
Traditionally, predictive models have focused on a single
structure for a small molecule and have ignored the role of the
receptor (in receptor-mediated scenarios). Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that the 3D geometries of that molecule be taken 
into account and, as far as possible, that the receptor also be taken
into account simultaneously. Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger and co-workers
\cite{Albuquerque:1998ys}.  Recent techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.

@@

At this point, ``electronic lab notebooks have replaced the
traditional paper lab book across the pharmaceutical industry,'' but
are not yet common in academia, because of their high
cost~\cite{WavingGoodbye}.  It is worth mentioning that chemistry is
traditionally a field where many people have been accustomed to huge
bookshelves of reactions and compounds. Chemists changed to online
ELNs not only to increase efficiency, but also because it allowed easy
collaboration with trusted colleagues. Third party paper catalogs and
academic publications proved to be inefficient, and even more
critically, they did not take compounds of trusted colleagues into
account. Overall the change of management to ELNs, similar to
``Enterprise 2.0,'' required delivering on the promise to increase
interconnectivity, and to improve collaboration opportunities.  The
leading players in this area include CambridgeSoft's E-Notebook and
Accelrys' Symyx Notebook, but there are at least 35 different
companies producing ELNs as of February 2011~\cite{ELNreview}, for use
in chemistry, biology, quality control, and other research domains.  Another
knowledge management tool specifically for chemists is the web-based
service Reaxys, founded in 2009, whose company goal is to provide ``a
single, fully integrated chemical workflow solution.''

@@

Due to the importance of cheminformatics to the pharmaceutical
industry in particular, numerous commercial vendors provide libraries,
applications and database cartridges to handle a wide variety of
needs including Accelrys, BioSolveIT, ChemAxon, Chemical Computing Group,
Daylight, Open Eye, Schrodinger, Tripos, and Xemistry. More recently
there has been a growth in the development of Open
Source cheminformatics software, and this presents opportunities for
the rapid development and implementation of novel algorithms which can
build on the existing Open Source ecosystem. In this article we do not debate the
arguments for and against Open Source; we focus on Open Source
tools simply because this ensures that any interested reader will be able to explore
cheminformatics problems with a minimum of hindrances.

@@

It is important to note that certain challenges faced by nearly all
cheminformatics toolkits stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality -- and edge cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
only derives from the chemical graph, but typically also from the 3D
geometry of the chemical. It is important to realize that although
chemical representations have increased in sophistication over the
years, many chiral molecules or non-standard bonding cases such as 
organometallic compounds cannot be satisfactorily handled by
current representation systems. Perhaps a more challenging aspect is
that the chemical representation of a molecule really needs to capture
non-static graphs, in order to take into account delocalisation and
the tautomeric phenomena molecules undergo in practice.

Of equal importance to the development of new algorithms is freely
available data on which to train and test new methods. As has been
noted, recently large structure and bioactivity data collections have
become available. As a result, this enables much more robust
validation of cheminformatics methodologies as well as large scale
\emph{benchmarking of algorithms}. The latter is especially relevant
for the plethora of data mining techniques that are employed in
cheminformatics. At this point, the nature of Open Data focuses
primarily on structure and activity data types. On the other hand,
there is a distinct lack of textual data (such as journal articles)
that are ``Open'' in nature. While PubMed abstracts serve as a proxy
for journal articles, text mining methods in cheminformatics are
hindered by not being able to mine the full text of many scientific
publications.  Interestingly, patent information is publicly
accessible (cf. Google Patents) and represents an excellent resource
to support these types of efforts.

%
\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the 19th
century and flowering in the 1960s with the advent of modern computing
technologies. While many key cheminformatics techniques have been
available in the literature since the 1960's, the bulk of
cheminformatics tools and implementations have been proprietary. In
addition, most data sources of chemical information have been
proprietary (usually with restrictive licensing policies) until
recently. There are many reasons for this, primarily revolving around
issues of profits, competitive intelligence and intellectual property,
which are relevant to the pharmaceutical as well as other chemical
industries. This is in contrast to much of bioinformatics, where the
data and tools have been freely available since the inception of the
field. One could argue that the outcomes from cheminformatics methods
and software have a more direct impact on profits, in terms of
identifying lead-like compounds and improving the properties of drug
candidates. In contrast, much bioinformatics software is employed in
upstream areas such as target identification (identifying the protein
that a small molecule binds to).  While certainly
important, this type of research is perhaps not as directly related to
possible profits as an actual small molecule.

Furthermore, chemical data (structure, activity information etc.) has
been harder to acquire and when done on large scales can be extremely
intensive in terms of time and effort, whereas bioinformatics data (sequences for
example) are much easier to obtain. While both fields have theoretical
components, it is possible that the free availability of large amounts
of data in bioinformatics drove the development of publicly
available tools to handle it. In contrast, the proprietary nature of
chemical data would imply that tools to process and analyze that data
were of interest primarily to the owners. As a result, the incentive
to make cheminformatics software public has been low and since the
users of it were primarily industrial, commercial software was the
norm.

>>>>

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: the
  interdisciplinary collaboration between computer scientists and
  molecular biologists that led as one of its major accomplishments to the
  sequencing of the human genome. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce
  new therapeutic molecules.  Only within the last decade -- as an
  effect of a change in mindsets, government-mandated policy changes,
  and, not least, as a result of chemists volunteering their time
  for an Open Science ``movement'' -- have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding therapies for rare and neglected diseases.
\end{abstract}

@@

Novel technologies in the life sciences churn out information at an
ever increasing rate -- public data stores such as that of the
European Bioinformatics Institute (EBI) contain in the order of 10
petabytes of biological information.  While biological information has
been publicly available for decades, the same has not been true for
chemical information, until recently. It was not until 2004 that a
large public small molecule structure repository (PubChem) became
freely available. This was followed by other databases. While much of
the foundational algorithms of cheminformatics have been described
since the 1950s, publicly available software implementing many of
these algorithms has only been accessible in the last ten to fifteen
years \cite{faulon2010}.

@@

\begin{figure}
\centering
\includegraphics[height=2in]{chemical-structures.png}
\caption{A hierarchy of chemical structure representations, increasing
  in information content from left to right and top to bottom. A
  chemist's traditional 2D depiction (A), a labeled graph (B), a 3D
  conformation (C) and a surface represention (D). All show differing
  (and valid) aspects of the same molecule and are suited for
  different purposes, with the choice of representation guided by its
  intended use.}
\label{figure:chemical-structures}
\end{figure}

@@

Even though some structure representations contain more explicit
information than others, they are all equally valid and are suited for
different problems. Fig.~\ref{figure:chemical-structures} shows 
a hierarchy of representations; when searching for
substructures, a 2D representation suffices, but when exploring 
protein binding, a 3D structure is vital.

@@

There is considerable progress in open chemical and bioactivity
databases like ChEMBL and PubChem, which are both freely
available and contain in the order of millions (ChEMBL) and tens of
millions (PubChem) of data points each. Integration of these disparate
data is essential for researchers to obtain the fullest possible 
perspective on what is presently known, tracking ongoing advances in 
science as these become available. Still, integration of data
across chemical databases remains a challenge, not only due
to the normalization problems in the chemical and bioactivity
landscape, but also due to sheer data volume. 
Given the requirement to ``make sense'' of as much chemical data 
as possible, we have to be able to search in complex data, for
example multi-labelled molecular graphs, where graph labels can change
from database to database. Hence, molecular graphs must be captured 
in a machine-readable fashion.

A fundamental challenge in cheminformatics is the unique
representation of chemical structures in a machine-readable format. A
(chemical) graph can be encoded in multiple ways, depending on how one
orders the nodes, resulting in multiple representations of the same
molecule. Considering protonation states 
(how the molecule changes when one proton is added
to it) and tautomer states (how the molecule can change when a proton
migrates from one part of the molecule to another), the challenge  
increases. The need for a unique representation that is invariant with
respect to the atom ordering arises due to the fact that graph
isomorphism (i.e., checking whether two structures represent the same molecule) is an
expensive operation. The first such ``unique representation
algorithm,'' or \emph{canonicalization} algorithm, was described by
Morgan \cite{Morgan1965}, allowing one to generate unique string
representations of chemical graphs letting one compare structures via
string comparisons. The SMILES (Simplified Molecular Input Line
Specification) format \cite{Weininger:1988kx} is an example of a
representation that can be canonicalized. Since the original
canonicalization algorithm was proprietary, multiple implementations
of the format have become available, each one employing a different
canonicalization algorithm, usually based on the Morgan algorithm. See
Warr~\cite{Warr:2011vn} for an extensive discussion on chemical
structure representations.

@@

\subsection{Molecules in their Biological Context}
\label{sec:prof-ident}

The preceding section focuses on the choice of representation of
chemical structures. In the quest for the discovery of novel
therapeutic agents, it is essential to study molecules as they are
situated in a biological context, underlying the processes of life, 
as metabolites, cofactors for enzymatic reactions, hormones, signalling
molecules and environmental xenobiotics. This gives rise to the need
to adequately represent and organise chemical data across biological
databases such as those containing pathways, protein information or 
biological samples. Integration and processing of data from such 
disparate domains underlies ``systems'' level biological research.
This gives rise to some additional challenges that standard chemical
structural representation cannot address: firstly, the representation of
\textit{classes of chemical entities} -- since, in many cases, a class of 
compounds behave in a certain context in a biological system rather than
a single molecule with a fully-specified structure -- and for similar
reasons, chemical concepts such as \textit{groups} (parts of molecules); 
secondly, the need to represent \textit{non-structural} groupings of 
compounds of interest, since compounds may bind to the same or similar 
targets or they may have  similar biological functions such as acting as 
an anti-neoplastic -- preventing the development of tumors. With the 
many different biological databases needing to refer to these different
chemical entities in an organised and standardized 
fashion to enable cross-database integration for a whole-systems 
perspective, \textit{ontologies} and other semantic technologies 
are being successfully employed to support annotation, classification and
semantic cross-domain querying of chemical entities within
biological systems. The most widely used ontology for biologically relevant
chemical entities is ChEBI \cite{chebi2010}, containing around 26,500 
entities in an interlinked semantic graph as of November 2011, and
being actively used for annotation in dozens of biological databases. 
Ontologies are based on logical languages of 
varying levels of expressivity, accompanied by sophisticated reasoning
algorithms. They can be used to construct various sorts of semantic
similarity (e.g. similarity in function, similarity in application) 
which complement traditional structural similarity methods \cite{couto2010}.
An open challenge is integrating graph-based chemical
structure information with the wider logic-based ontology information, which 
will allow combined automated reasoning and querying over the different domains 
encoded in a growing number of interlinked ontologies.% \cite{hastingsowled2010}. 
Enabling automated inferencing in such heterogenous linked systems brings us closer to
achieving a systems level understanding of bioactive chemistry \cite{Oprea:2007fk}.

@@

While molecular graph enumeration remains a challenge, 
an alternative to enumerating the space is to sample it
\cite{goldberg1999}. Sampling procedures based on Metropolis or
Genetic Algorithms have been developed and used to elucidate
compounds from NMR data. It should be noted that even in the
face of the complexity of molecular enumeration,
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB} (only
considering molecules up to 13 heavy atoms and using only carbon,
nitrogen, oxygen, sulfur and chlorine). Yet, the current enumeration
software products do not generally produce stereoisomers or tautomers;
these require specific enumeration procedures, which are still the
subject of research by the cheminformatics community.

Another relevant application of enumeration methods is the
\emph{generation of chemical reaction networks}. The problem here
consists of enumerating all the possible compounds that can be
produced by applying reaction rules to a set of initial molecules. By
reversing the reaction rules one can also find sets of starting
compounds necessary for producing a given target, known as
\emph{retrosynthesis}. Designing new drugs or chemicals, understanding
the kinetics of combustion and petroleum refining, studying the
dynamics of metabolic networks, applying metabolic engineering and
synthetic biology to produce heterologous compounds (compounds from
different species) in microorganisms, all involve enumeration of
reaction networks. As reviewed in Chapter 11 in \cite{faulon2010}
several network enumeration techniques have been developed. However,
these techniques generally suffer from a combinatorial explosion of
product compounds. One way to limit the number of compounds generated
is to simulate the dynamics of the network while it is being
constructed and remove compounds of low concentration. Following that
idea, methods have been developed based on the Gillespie Stochastic
Simulation Algorithm (SSA) to compute on-the-fly species
concentrations. Chemical reaction network enumeration and sampling is
an active field of research, particularly in the context of
metabolism, either to study biodegradation, or to propose metabolic
engineering strategies to biosynthesize compounds of commercial
interest. With metabolic network design, a difficulty is that in
addition to network generation based on reactions, one also needs to
verify that there are enzymatic events possible for enabling reactions
catalyses (enzymes must be present to reduce the
energy required for some reactions to take place). That additional
task requires encompassing both chemical structures and protein
sequences and the development of tools that are at the
interface between cheminformatics and bioinformatics.

@@

The basis of predictive modeling in cheminformatics is that the
biological activity of a molecule is a function of its chemical
structure. Together with the \emph{similar property principle}
\cite{Johnson:1990qf} (structurally similar molecules exhibit
similar properties), the goal of any modeling approach is to capture
and characterize correlations between structural features and
observed biological activities. Simultaneously, such approaches must
also describe the likelihood of error when using the
models for decision making.  A variety of approaches can be employed
to assess the error in (or conversely, the reliability or confidence of) a
prediction, ranging from statistical approaches to more
empirical approaches such as defining an \emph{applicability domain} -
the region of the input (i.e., chemical structures) that can be
reliably predicted, usually delineated by similarity
to the training set.

In many cases, the activity of a small molecule is due to its
interaction with a receptor. Traditionally, QSAR \cite{Hansch:1962vn,
 Free:1964ys} approaches do not take into account receptor features,
focusing only on small molecule features, and therefore lose valuable
information on ligand-receptor interactions. As a result, techniques
such as docking (which predicts how molecules fit together),
pharmacophore modeling (a 3D approach to capturing protein-ligand
interactions) and proteochemometric methods have been designed to take
into account both ligand and receptor structures. The last method is
an extension of statistical QSAR methods to simultaneously
model the receptor and ligand, as first reported by Lapinsh \textit{et
 al.}~\cite{lapinsh2001}.

@@

Both methods have advantages and disadvantages. The vectorial approach
requires one to identify a subset of \emph{relevant} (to the property
being modeled) descriptors: the feature selection problem 
well discussed in the data mining literature. A kernel approach
does not require feature selection, but one faces the problem of
evaluating a data set in a pairwise fashion, and one must
identify an appropriate kernel. This is an important challenge as the
kernel should be selected to satisfy \emph{Mercer's condition} (a
well-known mathematical property in machine learning that makes a set
of observations easier to make predictions about), and this is not
always possible with traditional cheminformatics based kernels (such
as a multiple common substructure based kernel). These challenges can
make kernel based methods prohibitive on very large datasets.

@@

One of the key challenges faced by predictive modeling is the fact
that small molecules are not static and do not exist in isolation. 
Traditionally, predictive models have focused on a single
structure for a small molecule and ignored the receptor. Yet, small molecules can
exist in multiple tautomeric forms and usually in multiple
conformations. As a result, enhancing the accuracy of predictions
ideally will require that the 3D geometries of that molecule be taken 
into account and, as far as possible, that the receptor also be considered. 
Multi-conformer modeling has been addressed in the
4D-QSAR methodology described by Hopfinger \textit{et al.}
\cite{Albuquerque:1998ys}.  Recent techniques such as
\emph{multiple-instance learning} could also be applied to the
multi-conformer problem.

@@

At this point, ``electronic lab notebooks have replaced the
traditional paper lab book across the pharmaceutical industry,'' but
are not yet common in academia, because of their high
cost~\cite{WavingGoodbye}.  It is worth mentioning that chemistry is
traditionally a field where many people have been accustomed to huge
bookshelves of reactions and compounds. Chemists changed to online
ELNs not only to increase efficiency, but also because it allowed easy
collaboration with trusted colleagues. Third party paper catalogs and
academic publications proved to be inefficient, and even more
critically, they did not take compounds of trusted colleagues into
account. Overall the change of management to ELNs, similar to
``Enterprise 2.0,'' required delivering on the promise to increase
interconnectivity, and to improve collaboration opportunities.  The
leading players in this area include CambridgeSoft's E-Notebook and
Accelrys' Symyx Notebook, but there are at least 35 different
companies producing ELNs as of February 2011~\cite{ELNreview}, for use
in chemistry, biology, quality control, and other research domains.  Another
knowledge management tool specifically for chemists is the web-based
service Reaxys, founded in 2009, which aims to provide ``a
single, fully integrated chemical workflow solution.''

@@

Due to the importance of cheminformatics to the pharmaceutical
industry, numerous commercial vendors provide libraries,
applications and database cartridges to handle a wide variety of
needs including Accelrys, BioSolveIT, Chem-Axon, Chemical Computing Group,
Daylight, Open Eye, Schrodinger, Tripos, and Xemistry. More recently
there has been a growth in the development of Open
Source cheminformatics software, and this presents opportunities for
the rapid development and implementation of novel algorithms which can
build on the existing Open Source ecosystem. In this article we do not debate the
arguments for and against Open Source; we focus on Open Source
tools because this ensures that an interested reader will be able to explore
cheminformatics problems with a minimum of hindrances.

@@

Certain challenges faced by nearly all
cheminformatics toolkits stem from the graph representation of
chemistry. This representation, while succinct and amenable to
computation, is only an approximation of reality -- and edge cases
abound. Existing areas of interest are enumeration of colored graphs,
taking into account symmetry, and symmetry detection itself, which not
only derives from the chemical graph, but typically also from the 3D
geometry of the chemical. It is important to realize that although
chemical representations have increased in sophistication over the
years, many chiral molecules or non-standard bonding cases such as 
organometallic compounds cannot be satisfactorily handled by
current representation systems. A more challenging aspect is
that the chemical representation of a molecule really needs to capture
non-static graphs, in order to take into account delocalisation and
the tautomeric phenomena molecules undergo in practice.

Of equal importance to the development of new algorithms is freely
available data on which to train and test new methods. As has been
noted, recently large structure and bioactivity data collections have
become available. As a result, this enables much more robust
validation of cheminformatics methodologies as well as large scale
\emph{benchmarking of algorithms}. The latter is especially relevant
for the plethora of data mining techniques employed in
cheminformatics. At this point, the nature of Open Data focuses
primarily on structure and activity data types. On the other hand,
there is a distinct lack of textual data (such as journal articles)
that are ``Open''. While PubMed abstracts serve as a proxy
for journal articles, text mining methods in cheminformatics are
hindered by not being able to mine the full text of many scientific
publications.  Interestingly, patent information is publicly
accessible (cf. Google Patents) and represents an excellent resource
to support these types of efforts.

%
\section{Conclusions}
\label{sec:conclusions}
Cheminformatics, the computer science of chemical discovery, is an
industrial and academic discipline with roots dating back to the 19th
century and flowering in the 1960s with the advent of modern computing
technologies. While many key cheminformatics techniques have been
available in the literature since the 1960's, the bulk of
cheminformatics tools and implementations have been proprietary. In
addition, most data sources of chemical information have been
proprietary (usually with restrictive licensing policies) until
recently. There are many reasons for this, primarily revolving around
issues of profits, competitive intelligence and intellectual property,
which are relevant to the pharmaceutical as well as other chemical
industries. This is in contrast to much of bioinformatics, where the
data and tools have been freely available since the inception of the
field. This disparity could be attributed to the fact that 
the outcomes from cheminformatics methods
and software have a more direct impact on profits, in terms of
identifying lead-like compounds and improving the properties of drug
candidates, while bioinformatics software is employed in
upstream areas such as target identification, 
perhaps not as directly related to
possible profits as an actual small molecule.

%Furthermore, chemical data (structure, activity information etc.) has
%been harder to acquire and when done on large scales can be extremely
%intensive in terms of time and effort, whereas bioinformatics data (sequences for
%example) are much easier to obtain. 
While both fields have theoretical
components, it is possible that the free availability of large amounts
of data in bioinformatics drove the development of publicly
available tools to handle it. In contrast, the proprietary nature of
chemical data would imply that tools to process and analyze that data
were of interest primarily to the owners. As a result, the incentive
to make cheminformatics software public has been low and since the
users of it were primarily industrial, commercial software was the
norm.
